---
title: "Homework 4"
author: "Arturo Alvarez"
date: "1, 15, 2026"
output:
  html_document: 
    theme: spacelab
---

## Instructions
Answer the following questions and/or complete the exercises in RMarkdown. Please embed all of your code and push the final work to your repository. Your report should be organized, clean, and run free from errors. Remember, you must remove the `#` for any included code chunks to run. Any plots must have appropriate titles and aesthetics.  

## Load the tidyverse
```{r}
library("tidyverse")
options(scipen = 999) #disable scientific notation
```

For this homework, we will be using the `msleep` dataset from the `ggplot2` package. This dataset contains information about the sleep habits of various mammals. [Reference](https://ggplot2.tidyverse.org/reference/msleep.html): V. M. Savage and G. B. West. A quantitative, theoretical framework for understanding mammalian sleep. Proceedings of the National Academy of Sciences, 104 (3):1051-1056, 2007.



```{r}
msleep <- msleep
```


**1. Use a summary function of your choice to get an idea of the structure of the msleep data.**

```{r}
summary(msleep)
```


**2. What are the names of the variables? To get a better idea of what the variables mean, you can use `?msleep`.**

#The name of the variables include name, 
```{r}
?msleep
names(msleep)
```


**3. Make a plot that shows the number of mammals in each vore type. Which vore type is most represented in the data?**

```{r}
ggplot(data=msleep,
       mapping=aes(x=vore))+
  geom_bar()+
  scale_x_discrete(na.translate=F)
```

**4. What is the average length of time that mammals sleeps?**
```{r}
mean(msleep$sleep_total)
```

**5. Use `summary()` to get a quick idea of the range of values for the variable `bodywt`.**
```{r}
summary(msleep)
```

**6. Build a plot that shows the range of `bodywt.` by vore type. Because the range of `bodywt` is so variable, you will need to scale the y-axis. You do this by adding `scale_y_log10()` as another layer.**
```{r}
ggplot(data=msleep, 
       mapping=aes(x=vore, y=bodywt))+
  scale_y_log10()+
geom_boxplot()+
  labs(title="Range of bodywt by vore type")
```

**7. Is there a relationship between body weight and total sleep time? Add appropriate labels and a title to your plot.**

#Based on the data, larger animals seem to sleep less as they grow more in body weight.
```{r}
ggplot(data=msleep, 
       mapping=aes(x=bodywt, y=sleep_total))+
  scale_x_log10()+
geom_point()+
  labs(title="Relationship between body weight and total sleep time")
```
  
**8. Which vore type tends to sleep the most? Build a plot to support your answer.**
#Based on the data, insectivores tend to sleep the most out of all vore types.
```{r}
ggplot(data=msleep, 
       mapping=aes(x=vore, y=sleep_total))+
  geom_boxplot()+
  labs(title="Total Sleep Time by Vore Type")
```

**9. Is there a relationship between brain weight and body weight? Build a plot to support your answer.**
#There does seem to be a relationship between brain weight and body weight. With a bigger brainweight resulting in a larger body weight and vice versa, although they don't seem to be proportional with eachother. 

```{r}
ggplot(data=msleep, 
       mapping=aes(x=brainwt, y=bodywt))+
  geom_point()+
  scale_x_log10()+
  scale_y_log10()+
  labs(title="Relationship between Brainwt and Bodywt")
```

**10. Build one plot of your choice that provides additional insight into the msleep dataset. Be sure to include appropriate labels and a title.**
#This data set provides extra insight on the relationship between the amount of rem sleep to ones total sleep, giving a positive relationship between the two. With more total sleep resulting in more time in rem sleep till eventually capping off at around 17 hours.

```{r}
ggplot(data=msleep, 
       mapping=aes(x=sleep_rem, y=sleep_total))+
  geom_point()+
  labs(title="Relationship between sleep_rem and sleep_total")
```

## Knit and Upload
Please knit your work as an .html file and upload to Canvas. Homework is due before the start of the next lab. No late work is accepted. Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  